Supporting Information to: Species sympatry shapes brain size evolution in Primates

### 2022-01-28 Article/Supplementary_material_forsubmission.Rmd
Update for submission - Separation Sup mat and Main text - Add Rmd word compilation

---
title: "Species sympatry shapes brain size evolution in Primates"
# date: "`r format(Sys.time(), '%B %d, %Y')`"

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: Benjamin Robira
    email: benjamin.robira@normalesup.org
    institute: [CEFE, MH]
    correspondence: true
  - name: Benoît Perez-Lamarque
    email: benoit.perez@ens.psl.eu
    institute: [ENS, MNHN]
institute:
  - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
  - MH: Anthropologie et Ethnobiologie, Centre National de la Recherche Scientifique/Muséum National d'Histoire Naturelle, University Paris Diderot, Sorbonne Paris Cité, Musée de l'Homme, Paris, France.
  - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France.
  - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France.

@@

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(include=FALSE, echo=FALSE, message=FALSE)
knitr::opts_chunk$set(dpi=300) # Figure resolution and size out.width = '100%', 
knitr::opts_chunk$set(fig.pos = 'p'#, fig.align = 'center'
                      ) # Places figures on pages separate from text, centered
knitr::opts_chunk$set(fig.env="figure") # Latex figure environment
knitr::opts_knit$set(eval.after = "fig.cap") #To insert R code into R figure caption
```

<!-- TC:ignore -->

@@

#Import own function
source("T:/Saved_PhD/Empirical_analysis/Scripts&Functions/Functions/toolbox.R", local = knitr::knit_global())

@@

load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversification_withautocorr.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdirectionSelection.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversificationAndSympatry.RData")
```

\captionsetup{list=no}
\newpage

```{r wordCount}
words <- RmdWords("Article.Rmd")
```

:::: {.graybox data-latex=""}

**Abstract |** The main hypotheses related to animal intelligence evolution highlight the role of conspecifics. Yet, space is often simultaneously occupied by species sharing the same ecological guild. These sympatric heterospecifics can compete for food, thereby stimulating or hampering cognition. Considering brain size as a proxy for cognition, we used primates to test for the intertwining between species sympatry and cognition. We retraced the evolutionary history of several brain areas with evolutionary models considering or not sympatry. Sympatry-related models best predicted the evolution of brain areas related to long-term memory of interactions with the social or ecological environment, with a decrease of their size the higher the sympatry. By contrast, the whole brain or brain areas used in immediate information processing were best described by models not considering sympatry. Moreover, sympatry negatively affected primate diversification. Overall, this comparative study suggests that species sympatry contributes to shaping primate cognition and diversification. We speculate that this is due to an over-complexification of resource spatio-temporality.

<!-- Reviewer possibles: Decasien, Powell, Barton, Kamilar, Harmon, Drury -->

\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Co-occurrence - Diversification - Frugivory - Primates - Sympatry

\faInfoCircle  \hspace{0.01in} **Word Count:** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->

::::

```{r sampleCalculation,  warning = FALSE, message = FALSE}
###Set working directory
setwd("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates")

repetition=2*2*2*10#length(frugivoryThresholdVector)*length(folivoryThresholdVector)*length(geographicThresholdVector)*randomSampling

checkSampleFruit <- rep(NA, times=repetition)
checkSampleLeaf <- rep(NA, times=repetition)
checkSampleRange <- rep(NA, times=repetition)
checkSampleBrain <-  rep(NA, times=repetition)
checkSampleEQ <-  rep(NA, times=repetition)
checkSampleNeocortex <-  rep(NA, times=repetition)
checkSampleHippocampus <- rep(NA, times=repetition)
checkSampleCerebellum <- rep(NA, times=repetition)
checkSampleStriatum <- rep(NA, times=repetition)
checkSampleMOB <- rep(NA, times=repetition)
checkSampleRange <- rep(NA, times=repetition)

counter=0
for(a in 1:2){
  for(b in 1:2){
    for(c in 1:2){
      for(d in 1:10){
        counter=counter+1
        tryCatch(
        {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleFruit",a,"_",b,"_",c,"_",d, ".txt", sep=""))
         checkSampleFruit[counter] <- toAdd[1]
        }, error=function(e){
          #Do nothing
        }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleLeaf",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleLeaf[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleRange",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleRange[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleBrain",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleBrain[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleEQ",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleEQ[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleNeocortex",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleNeocortex[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleHippocampus",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleHippocampus[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleCerebellum",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleCerebellum[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleStriatum",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleStriatum[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
        tryCatch(
          {toAdd <- read.delim(paste("Processed_data/Sample_size/checkSampleMOB",a,"_",b,"_",c,"_",d, ".txt", sep=""))
           checkSampleMOB[counter] <- toAdd[1]
          }, error=function(e){
            #Do nothing
          }
        )
      }
    }
  }
}

checkSampleFruit <- unlist(checkSampleFruit)
checkSampleLeaf <- unlist(checkSampleLeaf)
checkSampleRange <- unlist(checkSampleRange)
checkSampleBrain <-  unlist(checkSampleBrain)
checkSampleEQ <-  unlist(checkSampleEQ)
checkSampleNeocortex <-  unlist(checkSampleNeocortex)
checkSampleHippocampus <- unlist(checkSampleHippocampus)
checkSampleCerebellum <- unlist(checkSampleCerebellum)
checkSampleStriatum <- unlist(checkSampleStriatum)
checkSampleMOB <- unlist(checkSampleMOB)

#Min values
minFruit <- min(checkSampleFruit)
minLeaf <- min(checkSampleLeaf)
minRange <- min(checkSampleRange)
minBrain <- min(checkSampleBrain)
minEQ <- min(checkSampleEQ)
minNeocortex <- min(checkSampleNeocortex)
minHippocampus <- min(checkSampleHippocampus)
minCerebellum <- min(checkSampleCerebellum)
minStriatum <- min(checkSampleStriatum)
minMOB <- min(checkSampleMOB)

minAllAreas <- min(
minEQ,
minNeocortex,
minHippocampus,
minCerebellum,
minStriatum,
minMOB
)

#Max values
maxFruit <- max(checkSampleFruit)
maxLeaf <- max(checkSampleLeaf)
maxRange <- max(checkSampleRange)
maxBrain <- max(checkSampleBrain)
maxEQ <- max(checkSampleEQ)
maxNeocortex <- max(checkSampleNeocortex)
maxHippocampus <- max(checkSampleHippocampus)
maxCerebellum <- max(checkSampleCerebellum)
maxStriatum <- max(checkSampleStriatum)
maxMOB <- max(checkSampleMOB)

maxAllAreas <- max(
maxEQ,
maxNeocortex,
maxHippocampus,
maxCerebellum,
maxStriatum,
maxMOB
)

# Introduction

<!-- On the road to brain size evolution, generally considered as an equivalent of cognition evolution, mysteries are plenty [@van2006some; @dunbar2017there]. Overall, the evolution of the brain is the consequence of constraints (e.g. energetic limitations, @navarrete2011energetics), but also of socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference].  -->
<!-- Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for retracing human evolutionary history [@byrne2000evolution].  -->

<!-- \initial{\textcolor{black}{C}}ognition evolution results from the balance between socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference] and physiological and energetic constraints [@navarrete2011energetics]. Primates are pivotal species for cognitive studies [@byrne2000evolution] because their cognition is thought to be promoted by interactions of individuals with conspecifics within the social unit [@byrne1994machiavellian; @dunbar2017there], among generations [@wilson1991molecular; @whiten2007evolution; @reader2002social; @herrmann2007humans; @tomasello2009cultural; @van2011social], between social units [@ashton2020interactions], or with the rest of their environment [@clutton1980primates; @milton1981distribution; @rosati2017foraging]. However, space is a place often occupied by many species belonging to the same ecological guild and we can predict that such interactions with sympatric heterospecifics are also likely to strongly shape the evolution of cognition. -->

<!-- |   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities (but see discussion in @logan2018beyond). Yet, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic] and these areas are likely to be differently affected by species sympatry. For instance, the Main Olfactory Bulb (MOB) or the Hippocampus , home of a spatio-temporal memory [@burgess2002human], are largely involved in efficient foraging, especially for frugivorous primates, as fruits are the archetype of a hard-to-find resource yet predictable [@janmaat2016spatio; @robirabotany]: we therefore expect that primate species with larger MOB or Hippocampus areas, taken as equivalent to more advanced perceptive and cognitive abilities for efficient foraging, might be more fit than their sympatric species with smaller areas. Similarly, the Striatum underpins reward expectation and action, goal-directed behaviour and planning abilities [@johnson2007integrating] which is key when foraging. -->
<!-- are expected to be influenced by the ecological environment. The Striatum is also stimulated during social interactions [@baez2013role]: its size is thus expected to be positively influenced when contacts with other species increase, as in species performing mixed-species groups, which is nonetheless rather infrequent in primates (but see callitrichine primates in particular, @heymann2015unlike). Finally, besides foraging-related areas (Striatum, MOB or Hippocampus), more generally-used areas, such as the Cerebellum [@koziol2014consensus; @sokolov2017cerebellum] and the Neocortex [@wiltgen2004new] underlying movement and/or general information processing and retention, could also be stimulated by the presence of other species movement traces, noises or odors to decode. To sum up, a first possibility is that sympatric heterospecifics could altogether stimulate cognition and promote larger sizes of the brain areas related to interacting with the social or ecological environment.  -->

<!-- |   However, if the changes induced by sympatric species increases the environmental unpredictability too much, sympatry could on the contrary change the selective pressure applying to cognitive abilities [@grove2013evolution: @robirainreview]. Indeed, due to the cost of maintaining the brain functional [@raichle2006brain], it is expected that the increase in environmental unpredictability only stimulates cognition up to a certain threshold (i.e. positive selection for bigger brain), before smaller brain sizes of foraging-related areas become more adaptive (because costs are no longer compensated by the benefit of memory). Similarly, it is possible that processing heterospecifics cues, and other environmental cues (e.g. phenology cues) might not necessitate the same cognition level, with the former being potentially simpler to process (since relying on easily accessible social cues) than the latter (relying on trial-and-error processes and personal experience of potentially elusive environmental cues) [@laland2004social]. In this case, the size of brain areas involved in immediate information processing should be less large in sympatric than in non-sympatric conditions. To sum up, a second possibility is that sympatric heterospecifics could hamper the benefits of cognition on foraging, and thus slowdown brain size increase. In an extreme case, it could even induce positive selection for smaller brain size. -->

<!-- |    Here, we investigated the intertwine between species sympatry and cognition using frugivorous primates as study example. To infer the effect of species interactions on brain size evolution within frugivorous primates, we evaluated support for competitive or non-competitive evolutionary scenarios, accounting or not for sympatry, and investigated the directionality of the selection induced by sympatry history upon brain size evolution. Finally, we tested for correlative patterns between brain size or current sympatry with the evolutionary success (species diversification) in primates. -->

\initial{\textcolor{black}{C}}ognition evolution is shaped by the balance between socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference] and physiological and energetic constraints limiting them [@navarrete2011energetics]. Primates are pivotal species for cognitive studies [@byrne2000evolution] because their cognition is thought to be promoted by interactions of individuals with conspecifics within the social unit [@byrne1994machiavellian; @dunbar2017there], among generations [@wilson1991molecular; @whiten2007evolution; @reader2002social; @herrmann2007humans; @tomasello2009cultural; @van2011social], between social units [@ashton2020interactions], or with the rest of their environment [@clutton1980primates; @milton1981distribution; @rosati2017foraging]. However, space is often occupied by many species belonging to the same ecological guild. Because of competition for food, we can predict that indirect interactions with sympatric heterospecifics are also likely to strongly shape the evolution of cognition, because they either affect the resource landscape by depleting it (Hypothesis 1), or the landscape of usable “social” cues to locate available food ([@Avargues-Weber2013]; Hypothesis 2). In addition, direct interactions with heterospecifics, as much as with conspecifics [@dunbar2017there], might also affect cognition (Hypothesis 3).

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities, with larger size considered equivalent to more advanced cognitive abilities [@Benson-Amram2016]. Yet, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. These areas are likely to be under different selective pressures, and thus, to follow different evolutionary paths. In particular, they might be differently affected by species sympatry. First, brain areas involved in the storing of spatio-temporal information, such as the Hippocampus, home of an associative memory used for spatio-temporal navigation [@burgess2002human], should in particular be affected if the resource landscape is complexified by sympatric species foraging on the same resource (Hypothesis 1: memory is affected). Second, brain areas involved in processing more immediate sensory information, such as the Main Olfactory Bulb (MOB), Cerebellum [@koziol2014consensus; @sokolov2017cerebellum], and the Neocortex [@wiltgen2004new] should be particularly affected if the landscape of cues varies with sympatry (Hypothesis 2: cue processing is affected). Besides individual foraging, cognition can also be triggered by direct "social" interactions with other individuals [@byrne1994machiavellian; @dunbar2017there]. The Striatum is a brain area stimulated during social interactions [@baez2013role], and should, therefore, be affected by direct social interactions with heterospecifics (Hypothesis 3).

|   Under these (non-exclusive) hypotheses, sympatry could stimulate or hamper cognition evolution. Reasonable food depletion should promote memory, which stands as a valuable tool to infer food availability and location when food is rare and ephemeral but predictable [@milton1981distribution; @rosati2017foraging]. In this case, the size of the Hippocampus should be larger the higher the sympatry (Prediction 1.1). On the other hand, maintaining a functional brain is energetically costly [@raichle2006brain], and if changes induced by sympatric species increase environmental unpredictability too much, which would make cognitive foraging not advantageous anymore [@grove2013evolution; @robirainreview], the Hippocampus size should be smaller the higher the sympatry (Prediction 1.2). In the meanwhile, "social" cues  left out by heterospecifics might also add to environmental ones. If species sympatry increases the load of usable cues to locate available food, we should then expect larger sizes of the MOB, the Cerebellum, or the Neocortex. Interactions between species can also occur directly, with the formation of mixed-group species [@Goodale2010]. This, in particular, should promote larger Striatum size in sympatry.

|   Here, we investigated the intertwining between species sympatry and cognition using frugivorous primates as a study example. Frugivorous primates are an interesting group for such a question because fruits are the archetype of a hard-to-find resource yet predictable [@janmaat2016spatio], for which cognition thus considerably shapes the foraging strategy [@trapanese2019and]. To infer the effect of species sympatry on brain size evolution within frugivorous primates, we evaluated the support for evolutionary models accounting or not for sympatry, and investigated the directionality of the selection induced by sympatry history upon brain size evolution. Finally, we tested for correlative patterns between brain size or current sympatry and the evolutionary consequences (assumed to be reflected by species diversification) in all primates.

# Results

```{r calculationValueForFirstResults}
#Import data
summaryDataForPlot <- read_delim("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/OutputEvolModel/Dataplot.txt","\t", escape_double = FALSE, trim_ws = TRUE)

summaryDataForPlot <- summaryDataForPlot[!is.na(summaryDataForPlot$geographicCode),]
# summaryDataForPlot <- summaryDataForPlot[summaryDataForPlot$DietaryGuild=="Fruit",]

summaryDataForPlot$Family <- Data_powell2$MSW05_Family[match(summaryDataForPlot$Species_abbrv,Data_powell2$Species_abbrv)]#summaryData$Family[match(summaryDataForPlot$Species_abbrv, summaryData$Species_abbrv)]

summaryDataForPlot[is.na(summaryDataForPlot$Family ),c(1,2,3)]
#summaryDataForPlot$SpeciesForPhylogeny[which(is.na(summaryDataForPlot$Family))]
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Galagoides_demi"]<-"Galagonidae"
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Macaca_munz"]<-"Cercopithecidae"
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Microcebus_mitt"]<-"Cheirogaleidae"
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Mirza_zaza"]<-"Lorisidae"
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Nycticebus_java"]<-"Galagonidae"
summaryDataForPlot$Family[summaryDataForPlot$Species_abbrv=="Tarsius_lari"]<-"Tarsiidae"

lemuriformes.v <- 
c(  
"Daubentoniidae",
"Lemuridae",
"Indriidae",
"Cheirogaleidae",
"Lepilemuridae"
)

platyrrhini.v <- c(
"Atelidae",   
"Aotidae",
"Pitheciidae",
"Cebidae"
)

meanseMOBlemu <- round(c(
  mean(
    (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% lemuriformes.v], 
    na.rm=TRUE), 
  sd(
    (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% lemuriformes.v],
    na.rm=TRUE)/
    sqrt(
      length(
        (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% lemuriformes.v &
                                                               !is.na(summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)]))
), digit=2)

meanseMOBother <- round(c(
  mean(
    (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% lemuriformes.v], 
    na.rm=TRUE), 
  sd(
    (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% lemuriformes.v],
    na.rm=TRUE)/
    sqrt(
      length(
        (summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% lemuriformes.v &
                                                               !is.na(summaryDataForPlot$MOB/summaryDataForPlot$Bodymass)]))
), digit=2)

meanseStriatumplaty <- round(c(
  mean(
    (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% platyrrhini.v], 
    na.rm=TRUE), 
  sd(
    (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% platyrrhini.v],
    na.rm=TRUE)/
    sqrt(
      length(
        (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %in% platyrrhini.v &
                                                               !is.na(summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)]))
), digit=2)

meanseStriatumother <- round(c(
  mean(
    (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% platyrrhini.v], 
    na.rm=TRUE), 
  sd(
    (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% platyrrhini.v],
    na.rm=TRUE)/
    sqrt(
      length(
        (summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)[summaryDataForPlot$Family %nin% platyrrhini.v &
                                                               !is.na(summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass)]))
), digit=2)

# summaryDataForPlot[order(summaryDataForPlot$MOB/summaryDataForPlot$Bodymass, decreasing = TRUE),]
# summaryDataForPlot[order(summaryDataForPlot$Hippocampus/summaryDataForPlot$Bodymass, decreasing = TRUE),]
# summaryDataForPlot[order(summaryDataForPlot$Striatum/summaryDataForPlot$Bodymass, decreasing = TRUE),]

We gathered a large database on the whole brain size, the size of the MOB, Cerebellum, and the Neocortex, involved in immediate sensory information processing, and the Striatum and the Hippocampus, related to either social and foraging-related memory, containing between `r minAllAreas` to `r maxEQ` frugivorous primate species (depending on the brain area considered). After pondering by whole-body mass, we observed ample variations in brain area relative sizes. For instance, the lemuriformes, that are known to prioritize smell compared to other primate species, have the largest relative MOB size (Lemuriformes: mean $\pm$ SE = `r meanseMOBlemu[1]` $\pm$ `r meanseMOBlemu[2]`, other: `r meanseMOBother[1]` $\pm$ `r meanseMOBother[2]`, \@ref(fig:figbrain)). Similarly, platyrrhini, and callitrichine primates in particular, are known to form poly-specific associations [@heymann2000behavioural]. The latter show the highest relative size of the Striatum (Platyrrhini: mean $\pm$ SE = `r meanseStriatumplaty[1]` $\pm$ `r meanseStriatumplaty[2]`, other: `r meanseStriatumother[1]` $\pm$ `r meanseStriatumother[2]`, \@ref(fig:figbrain)).

|   To get the evolutionary history of species sympatry between frugivorous lineages, we first reconstructed primate biogeography history (N$_{species}$ = `r minRange`; [@matzke2013probabilistic; @matzke2016stochastic]) when considering `r length(areaName)` biogeographic areas (Figure \@ref(fig:figmap); [@kamilar2009environmental]) and their diet evolution (N$_{species}$ = `r minFruit + minLeaf` to `r maxFruit + maxLeaf`; [@bollback2006simmap]). We then calculated the likelihoods of models considering the role of species sympatry in the evolution of either the whole brain (using the encephalic quotient, EQ), or the size of the aforementioned specific brain areas relative to the whole-body mass (Figure. \@ref(fig:figbrain)). We specifically considered the matching competition (MC) model [@nuismer2015predicting] and density dependence models (DD$_{lin}$ and DD$_{exp}$), which assume a linear or exponential dependence of the brain sizes evolutionary rates on the number of sympatric lineages ([@drury2016estimating] see [Models of trait evolution: does species sympatry shape brain size evolution?]). We compared the support of models considering species sympatry to the support of simpler models assuming no effect of species sympatry, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) considering that traits are constrained around an optimal value (e.g. stabilizing selection; see [@blomberg2020beyond] for a review), or the Early-Burst model (EB, [@blomberg2003testing]), the latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density dependence is not an artefact due to time dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weights of the Akaike Information Criteria corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does species sympatry shape brain size evolution?]).

|   We found that models not considering species sympatry were the most likely in describing the evolutionary history of the EQ, the Neocortex, and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in immediate sensory information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum], and also in memory consolidation for the Neocortex [@wiltgen2004new]. The fact that these biggest areas are best described by the Ornstein-Uhlenbeck process suggests a stabilization towards an optimal size, which may illustrate the trade-offs between costs and benefits of brain development [@isler2009expensive]. By contrast, density-dependence models considering sympatry were best supported in the foraging-related and social-related areas respectively: the Hippocampus, specialised in spatio-temporal memory [@burgess2002human] and the Striatum, involved in social interactions [@baez2013role]. The fact that we inferred positive rates *r* of density-dependence (Figure \@ref(fig:figresultsevolution)), suggested an acceleration of the evolutionary tempo of trait evolution together with increased diversity of sympatric lineages for the Hippocampus and the Striatum. The Main Olfactory Bulb (MOB), the area involved in sensory abilities, also tended to be best fitted by models considering sympatry as a whole: yet, Brownian-Motion was as likely as density-dependent or MC models, preventing firm conclusions on whether sympatry affected or not MOB size evolution (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)).

```{r samplecooccurrence}

minNspeciesCoocc <- round(min(dataRangePrimate$Number_species_cooccurrence, na.rm=TRUE), digit=2)
maxNspeciesCoocc <- round(max(dataRangePrimate$Number_species_cooccurrence, na.rm=TRUE), digit=2)
meanNspeciesCoocc <- round(mean(dataRangePrimate$Number_species_cooccurrence, na.rm=TRUE), digit=2)
seNspeciesCoocc <- round(sd(dataRangePrimate$Number_species_cooccurrence, na.rm=TRUE)/sqrt(length(dataRangePrimate$Number_species_cooccurrence[!is.na(dataRangePrimate$Number_species_cooccurrence)])), digit=2)

whichSpeciesMinNCoocc <- paste(sort(dataRangePrimate$Species[which(round(dataRangePrimate$Number_species_cooccurrence, digit=2)==minNspeciesCoocc)]), collapse=", ")
whichSpeciesMaxNCoocc <- paste(sort(dataRangePrimate$Species[which(round(dataRangePrimate$Number_species_cooccurrence, digit=2)==maxNspeciesCoocc)]), collapse=", ")
whichSpeciesMinNCoocc <- gsub("_", " ", whichSpeciesMinNCoocc)
whichSpeciesMaxNCoocc <- gsub("_", " ", whichSpeciesMaxNCoocc)

minOverlapSpecies <- round(min(dataRangePrimate$Overlap_average, na.rm=TRUE), digit=2)
maxOverlapSpecies <- round(max(dataRangePrimate$Overlap_average, na.rm=TRUE), digit=2)
meanOverlap <- round(mean(dataRangePrimate$Overlap_average, na.rm=TRUE), digit=2)
seOverlap <- round(sd(dataRangePrimate$Overlap_average, na.rm=TRUE)/sqrt(length(dataRangePrimate$Overlap_average[!is.na(dataRangePrimate$Overlap_average)])), digit=2)

whichSpeciesMinOverlap <- paste(dataRangePrimate$Species[which(round(dataRangePrimate$Overlap_average, digit=2)==minOverlapSpecies)], collapse=", ")
whichSpeciesMaxOverlap <- paste(dataRangePrimate$Species[which(round(dataRangePrimate$Overlap_average, digit=2)==maxOverlapSpecies)], collapse=", ")
whichSpeciesMinOverlap <- gsub("_", " ", whichSpeciesMinOverlap)
whichSpeciesMaxOverlap <- gsub("_", " ", whichSpeciesMaxOverlap)

```{r estimateRegGradient, error=TRUE, include=TRUE}

#Hippocampus
model <- get(paste("modelBrain", traitName[3], sep="_"))

estimateGradientHippocampus <- textEstOneVar(summary(model)$coefficients[2,c(1,4,5)])
testGradientHippocampus <- textTestOneVar(summary(model)$coefficients[2,c(3,6)], statistics="")

#Striatum
model <- get(paste("modelBrain", traitName[6], sep="_"))

estimateGradientStriatum <- textEstOneVar(summary(model)$coefficients[2,c(1,4,5)])
testGradientStriatum <- textTestOneVar(summary(model)$coefficients[2,c(3,6)], statistics="")

#MOB
model <- get(paste("modelBrain", traitName[7], sep="_"))

estimateGradientMOB <- textEstOneVar(summary(model)$coefficients[2,c(1,4,5)])
testGradientMOB <- textTestOneVar(summary(model)$coefficients[2,c(3,6)], statistics="")
```

|   Next, to understand the effect of sympatry on brain sizes (i.e. toward “bigger” or “smaller” brains the more sympatric species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered as covariates the average percentage of overlapping range with other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire distribution range based on IUCN data [@IUCN]. On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). In terms of the distribution range, the considered primate species co-occurred on average with `r meanNspeciesCoocc` other primate species ($\pm$ `r seNspeciesCoocc`), ranging from `r minNspeciesCoocc` other species to `r maxNspeciesCoocc`. The number of sympatric species never influenced significantly the relative size of the brain or other specific areas (Table \@ref(tab:tableRegGradient)). Conversely, we found that the average percentage of overlapping range correlated with the relative size of areas that were better fit with models considering sympatry: the Hippocampus and the Striatum (Hippocampus: $t$`r testGradientHippocampus`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`; Striatum: `r estimateGradientStriatum`), which means that higher range overlap between frugivorous species associates with lower relative size, insensitive to data and phylogenetic uncertainty (Table \@ref(tab:tabledfsensitivity), Figure \@ref(fig:forestPlot)). Given the acceleration of the evolutionary tempo with species sympatry ($r>0$ in the density-dependence models), it suggests that compared with isolated species, sympatric species are subject to a positive selection towards smaller brains, and not to a less intense selection for advanced cognitive abilities.

```{r estimateRegDiversification, error=TRUE, include=TRUE}

>>>>

---
title: "Supporting Information to: Species sympatry shapes brain size evolution in Primates"
# date: "`r format(Sys.time(), '%B %d, %Y')`"

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author:
#   - name: Benjamin Robira
#     email: benjamin.robira@normalesup.org
#     institute: [CEFE, MH]
#     correspondence: true
#   - name: Benoît Perez-Lamarque
#     email: benoit.perez@ens.psl.eu
#     institute: [ENS, MNHN]
# institute:
#   - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
#   - MH: Eco-anthropologie (EA), Muséum National d'Histoire Naturelle, CNRS, Université de Paris, Musée de l'Homme, Paris, France.
#   - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France.
#   - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France.

@@

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(include=FALSE, echo=FALSE, message=FALSE)
knitr::opts_chunk$set(dpi=300) # Fig. resolution and size out.width = '100%', 
knitr::opts_chunk$set(fig.pos = 'p'#, fig.align = 'center'
                      ) # Places figures on pages separate from text, centered
knitr::opts_chunk$set(fig.env="figure") # Latex Fig. environment
knitr::opts_knit$set(eval.after = "fig.cap") #To insert R code into R Fig. caption
```

@@

library(tidyverse)
#Import own function
source("T:/Saved_PhD/Empirical_analysis/Scripts&Functions/Functions/toolbox.R", local = knitr::knit_global())

@@

load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversification_withautocorr.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdirectionSelection.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversificationAndSympatry.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/RDataforAppendix.RData")
```

# Data variability
